2.1 APL Data Quality Check with additonal cleaning

### 2023-12-04 2_scripts/2.1 APL Data Quality Check.qmd
APL Figures

## Load the APL data with encrypted ID variables
```{r}
load(file = "~/APL/RFD4682e.RData")
```

## Load paks
```{r}
library(dplyr)
```

@@

## Records are not patients, how many individual has been included?
```{r}
## ULI_ID: Unique Lifetime Identifier (ULI), numerical values of Alberta PHN.
length(unique(RFD4682_e$ULI_IDe))     # Unique = 180821
table(is.na(RFD4682_e$ULI_IDe))       # NA = 11298

@@

str(tab_SERO_ID)
tab_SERO_ID.dat<-as.data.frame(tab_SERO_ID)
tab_SERO_ID.dat[tab_SERO_ID.dat$Freq==2,]
```
## Why unique PERSON_ID and ULI_ID are not matching with each other?
```{r}
## Check the discrepancies between those two

@@

## Scenario Examples Showcase
```{r}
# Example with each Scenarios
RFD4682_3[RFD4682_3$Freq_ULI==6 & RFD4682_3$Freq_PER==5 ,]     # Scenario 5
RFD4682_3[RFD4682_3$Freq_ULI==15 & RFD4682_3$Freq_PER==15 ,]   # Scenario 1
RFD4682_3[RFD4682_3$Freq_ULI==2 & RFD4682_3$Freq_PER==0 ,]     # Scenario 2
RFD4682_3[RFD4682_3$Freq_ULI==0 & RFD4682_3$Freq_PER==5 ,]     # Scenario 3
RFD4682_3[RFD4682_3$Freq_ULI==1 & RFD4682_3$Freq_PER==2 ,]     # Scenario 4
RFD4682_3[RFD4682_3$Freq_ULI==0 & RFD4682_3$Freq_PER==0 ,]     # Scenario 0
```

## Add clean_IDe and Scenario back to dataset (encrypted)
```{r}
RFD4682_e<-merge(RFD4682_e, select(RFD4682_3, order_ID, Scenario, clean_IDe),
                      by = 'order_ID', all = TRUE)
```

#### ULI ID and PERSON_ID, in NAAT+ subset
```{r}
## ULI_ID: Unique Lifetime Identifier (ULI), numerical values of Alberta PHN.
length(unique(RFD4682_NAAT$ULI_ID))     # N = 5933
table(is.na(RFD4682_NAAT$ULI_ID))       # NA = 0

@@

## Check N=2865 cases with missing in both ULI and Person_ID
```{r}
RFD4682_e[RFD4682_e$Scenario==0,]
## Surprisingly, most of the records in this group has some useful variables,
## such as: age, gender and Anti_RBD test results

@@

## The Whole APL dataset
```{r}
# The NAAT status variable
table(RFD4682_e$NAAT_STATUS, useNA = 'ifany')

@@

# Cross-tab of those two
table(RFD4682_e$NAAT_STATUS, RFD4682_e$VACCINATION_STATUS, useNA = 'ifany')
```
## Categorize the time since last NAAT test
```{r}
hist(RFD4682_e$DAYS_SINCE_NAAT, breaks=20)

@@

## Vaccination Status
```{r}
table(RFD4682_NAAT$VACCINATION_STATUS, useNA = 'ifany')
hist(RFD4682_NAAT$DAYS_SINCE_IMM)

@@

>>>>

## Load the APL data with encrypted ID variables

```{r}
load(file = "../1_data/private/RFD4682e.RData")
```

## Load paks

```{r}
library(dplyr)         # Data management
library(lubridate)     # Manage variables in Date formats
library(zoo)           # for yearmonth Date format
library(reshape2)      # for melt() function
```

@@

## Records are not patients, how many individual has been included?

```{r}
## ULI_ID: Unique Lifetime Identifier (ULI), numerical values of Alberta PHN.
length(unique(RFD4682_e$ULI_IDe))     # Unique = 180821
table(is.na(RFD4682_e$ULI_IDe))       # NA = 11298

@@

str(tab_SERO_ID)
tab_SERO_ID.dat<-as.data.frame(tab_SERO_ID)
tab_SERO_ID.dat[tab_SERO_ID.dat$Freq==2,]
```

## Why unique PERSON_ID and ULI_ID are not matching with each other?

```{r}
## Check the discrepancies between those two

@@

## Scenario Examples Showcase

```{r}
# Example with each Scenarios
RFD4682_3[RFD4682_3$Freq_ULI==6 & RFD4682_3$Freq_PER==5 ,]     # Scenario 5
RFD4682_3[RFD4682_3$Freq_ULI==15 & RFD4682_3$Freq_PER==15 ,]   # Scenario 1
RFD4682_3[RFD4682_3$Freq_ULI==2 & RFD4682_3$Freq_PER==0 ,]     # Scenario 2
RFD4682_3[RFD4682_3$Freq_ULI==0 & RFD4682_3$Freq_PER==5 ,]     # Scenario 3
RFD4682_3[RFD4682_3$Freq_ULI==1 & RFD4682_3$Freq_PER==2 ,]     # Scenario 4
RFD4682_3[RFD4682_3$Freq_ULI==0 & RFD4682_3$Freq_PER==0 ,]     # Scenario 0
```

## Add clean_IDe and Scenario back to dataset (encrypted)

```{r}
RFD4682_e<-merge(RFD4682_e, select(RFD4682_3, order_ID, Scenario, clean_IDe),
                      by = 'order_ID', all = TRUE)
```

#### ULI ID and PERSON_ID, in NAAT+ subset

```{r}
## ULI_ID: Unique Lifetime Identifier (ULI), numerical values of Alberta PHN.
length(unique(RFD4682_NAAT$ULI_ID))     # N = 5933
table(is.na(RFD4682_NAAT$ULI_ID))       # NA = 0

@@

## Check N=2865 cases with missing in both ULI and Person_ID

```{r}
RFD4682_e[RFD4682_e$Scenario==0,]
## Surprisingly, most of the records in this group has some useful variables,
## such as: age, gender and Anti_RBD test results

@@

## The Whole APL dataset

```{r}
# The NAAT status variable
table(RFD4682_e$NAAT_STATUS, useNA = 'ifany')

@@

# Cross-tab of those two
table(RFD4682_e$NAAT_STATUS, RFD4682_e$VACCINATION_STATUS, useNA = 'ifany')
```

## Categorize the time since last NAAT test

```{r}
hist(RFD4682_e$DAYS_SINCE_NAAT, breaks=20)

@@

## Vaccination Status

```{r}
table(RFD4682_NAAT$VACCINATION_STATUS, useNA = 'ifany')
hist(RFD4682_NAAT$DAYS_SINCE_IMM)

@@

Estimation on number of individuals, using cleaned version of IDe

```{r}
# Total num of individual in the data:
table(is.na(unique(RFD4682_e$clean_IDe)), useNA = 'ifany')    # N=188,722, plus 1 NA

# Number of participants with available Anti-N results:
antin_na<-is.na(unique(RFD4682_e[is.na(RFD4682_e$clean_IDe)==F &
                             is.na(RFD4682_e$`N-IgG_RESULT`)==F,
                             ]$clean_IDe))
table(antin_na, useNA = 'ifany')  # N=39450

# Number of participants with available Anti-S results:
antis_na<-is.na(unique(RFD4682_e[is.na(RFD4682_e$clean_IDe)==F &
                             is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F,
                             ]$clean_IDe))
table(antin_na, useNA = 'ifany') # N=188,688

table(is.na(RFD4682_e$`N-IgG_RESULT`)==F,
      is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F, useNA = 'ifany')
```
Some additional cleaning/editing, making anti-S quantitative
```{r}
# Missing values
summary(is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F) 
# N=36, beware of "not available" values, better recode those as NAs

# Remove all the numbers, see what left
S_chr<-RFD4682_e$`RBD-IgGII_RESULT` %>%
  gsub("[[:digit:]]+", "",.)
table(S_chr, useNA = 'ifany')
## As we can see, the use of units in this data is quite messy, needs additonal cleaning

# How many of them use unit "AU/ml" (arbitury unit)?
AU<-RFD4682_e$`RBD-IgGII_RESULT` %>%
  grepl("ml", .,fixed = F, ignore.case = T)
table(AU) # N=109798

# index
index<-RFD4682_e$`RBD-IgGII_RESULT` %>%
  grepl("index", .,fixed = F, ignore.case = T)
table(index) # N=104946
```
#### Yuan found this: anti-N is only available for the Omicron stage
```{r}
# The APL Data:
RFD4682_e$COLLECTION_DATE %>%
  summary()

# Anti-N assay results is only available to N=39,450 participants
RFD4682_e[is.na(RFD4682_e$clean_IDe)==F &
          is.na(RFD4682_e$`N-IgG_RESULT`)==F,]$COLLECTION_DATE %>%
  summary()
# Anti-N: 2021/12/14 -- 2022/10/14

# Anti-S assay results is available to most of the participants
RFD4682_e[is.na(RFD4682_e$clean_IDe)==F &
          is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F,]$COLLECTION_DATE %>%
  summary()
```
#### A quick clean of the Anti-N data
```{r}
# Remove unecessary 'index'
table(RFD4682_e$`N-IgG_RESULT`)

RFD4682_e$N<-RFD4682_e$`N-IgG_RESULT` %>%
  gsub("Index", "",.) %>%
  as.numeric()

hist(RFD4682_e$N, breaks=25)

# Results
table(RFD4682_e$`N-IgG_INTERP`, useNA = 'ifany')
```
#### Calculate positivity directly

```{r}
# Create month var
RFD4682_e$Col_Date<-RFD4682_e$COLLECTION_DATE %>%
  as.Date(., tz="UTC") # a short time var, date only

library(lubridate)
library(zoo)

# Only keep the month, remove the data
RFD4682_e$mois <- month(RFD4682_e$Col_Date)
RFD4682_e$annee <- year(RFD4682_e$Col_Date)
RFD4682_e$Col_Month<-paste(RFD4682_e$annee, RFD4682_e$mois, 
                           sep = "-")
table(RFD4682_e$Col_Month, useNA='ifany')
# as char variable if needed for calculation, use as.yearmon() from {zoo} or just use the Col_Date var
```

```{r}
# Calculate positivity for each month
tb1<-RFD4682_e[is.na(RFD4682_e$`N-IgG_RESULT`)==F,] %>%
  group_by(Col_Month) %>%
  summarise(n = n())

tb2<-RFD4682_e[RFD4682_e$`N-IgG_INTERP`=="Positive",] %>%
  group_by(Col_Month) %>%
  summarise(n.pos = n())

tb1<-merge(tb1, tb2, by = "Col_Month", all.x = T, all.y = F)
tb1$prop.n<-tb1$n.pos/tb1$n
tb1$Col_Month<-as.yearmon(tb1$Col_Month)
tb1<-tb1[order(tb1$Col_Month),]

# Add R-G adjusted prevalances
# Sen=0.881, Spec = 0.947
# Source:
tb1$prop.n_RG<- (tb1$prop.n+0.881-1)/(0.881+0.947-1)

# remove NAs
tb1<-tb1[!is.na(tb1$Col_Month),]

tb1 %>%
  with(plot(y=.$prop.n, x = .$Col_Month, type="b",
            ylab="positivity", xlab="Month",col="blue")
       )
```

#### Apply ggplot, plot two together
```{r}
tb1_melt<-melt(select(tb1, Col_Month, prop.n, prop.n_RG), id.vars = "Col_Month")

tb1_melt %>%
  ggplot(aes(x=Col_Month, y=value, colour=variable)) +
  geom_point(aes(colour=variable)) +
  geom_line(aes(colour=variable))  +
  theme_bw()
```

```{r}

plot(x=temp$Time ,y=temp$Positivity, type = 'b', 
     col="blue", xlab="Time: 2022", ylab="Test Positivity",
     main="Anti-N Assay Positivity in Omicron Wave, using APL Data")
plot(x=temp$Time ,y=temp$Positivity_RG, 
     type = 'b', col="red")
```
